Kindly Bent to Free Us

### 2020-05-17 kindly/lang.tex
more comments by Rev#A processed

A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive/affine
($\MBORROW$) or shared/unrestricted ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match operates on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
%
We consider four primitive operations to manipulate resources:
$\create$, $\observe$, $\update$ and $\destroy$.
They serve a prototypes to demonstrate the typing and handling of
resources. For a concrete type of resource, there are further
arguments and perhaps different versions of the operations. But the
typing and behavior of the operations is analogous to the prototype
operations.

@@

\begin{figure*}[tp]
  % \begin{minipage}{0.7\linewidth}
    \begin{mathpar}
      \ruleSDVar
      \and
      \ruleSDApp
      \and
      \ruleSDRegion
      \and
      \ruleSDBorrow
      \and
      \ruleSDLam
      \and
      \inferrule[BorrowBinding]{
        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
        \BORROW\in\left\{\kun,\kaff\right\}
      }{
        \lregion{C}{}
        {\svar[\BORROW]{x}{\schm}^n}
        {\bbvar x k \schm}
        % {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\schm}}}
      }
      % \and
      % \ruleSDPair
      % \and
      % \ruleSDMatchPair
    \end{mathpar}
    \vspace{-10pt}
    \caption{Selected typing rules ($\inferS{C}{\E}{e}{\tau}$)
      and borrowing rules ($\lregion{C}{x}{\E}{\E'}$)}
    \label{selectrules:borrow}
    \label{selectrules:binders}
    \label{sdtyp:app}
    \label{selectrules:region}
    \label{env:rule:borrow}
    \vspace{-5pt}
  % \end{minipage}
\end{figure*}

@@

To demonstrate how \lang{} deals with resources, we introduce an abstract
type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be
unrestricted ($\kun_0$) and which is equipped with the four
operations described in \cref{motivation}: 
\begin{itemize}[topsep=0pt]
\item
$\create$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\tvar \tarr{} \tapp\tres\tvar}$
% Creates a new resource.
\item
$\observe$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$
% Reads from a shared borrow of a resource.
\item
$\update$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$
% Writes to an exclusive borrow of a resource.
\item
$\destroy$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\tapp\tres\tvar \tarr{} \tunit$
% Destroys a resource.
\end{itemize}

@@

>>>>

A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive/affine
($\MBORROW$) or shared/unrestricted ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match operates on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
%
We consider four primitive operations to manipulate resources:
$\create$, $\observe$, $\update$ and $\destroy$.
They serve a prototypes to demonstrate the typing and handling of
resources. For a concrete type of resource, there are further
arguments and perhaps different versions of the operations. But the
typing and behavior of the operations is analogous to the prototype
operations. Moreover, $\observe$ and $\update$ serve as eliminators
for borrow types.

@@

\begin{figure*}[tp]
  % \begin{minipage}{0.7\linewidth}
    \begin{mathpar}
      \ruleSDVar
      \and
      \ruleSDApp
      \and
      \ruleSDRegion
      \and
      \ruleSDBorrow
      \and
      \ruleSDReBorrow
      \and
      \ruleSDLam
      \and
      \inferrule[BorrowBinding]{
        \entail{C}{(\BORROW_n\lk k) \wedge (k \lk \BORROW_\infty)}\\
        \BORROW\in\left\{\kun,\kaff\right\}
      }{
        \lregion{C}{}
        {\svar[\BORROW]{x}{\schm}^n}
        {\bbvar x k \schm}
        % {\bvar{\borrow[\BORROW]{x}}{\borrowty[\BORROW] k{\schm}}}
      }
      % \and
      % \ruleSDPair
      % \and
      % \ruleSDMatchPair
    \end{mathpar}
    \vspace{-10pt}
    \caption{Selected typing rules ($\inferS{C}{\E}{e}{\tau}$)
      and borrowing rules ($\lregion{C}{x}{\E}{\E'}$)}
    \label{selectrules:borrow}
    \label{selectrules:binders}
    \label{sdtyp:app}
    \label{selectrules:region}
    \label{env:rule:borrow}
    \vspace{-5pt}
  % \end{minipage}
\end{figure*}

@@

To demonstrate how \lang{} deals with resources, we introduce an abstract
type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be
unrestricted ($\kun_0$) and which is equipped with the four
operations introduced in \cref{syntax}: 
\begin{itemize}[topsep=0pt]
\item
$\create$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\tvar \tarr{} \tapp\tres\tvar}$
% Creates a new resource.
\item
$\observe$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$
% Reads from a shared borrow of a resource.
\item
$\update$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$
% Writes to an exclusive borrow of a resource.
\item
$\destroy$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\tapp\tres\tvar \tarr{} \tunit$
% Destroys a resource.
\end{itemize}

@@

It is straightforward to give a standard semantics for \lang, but such
a semantics would not be very informative. In this section, we give a
big-step semantics that performs explicit bookkeeping of the number of
times a value is used and of the mode in which a reference to a
resource is used (e.g., borrowed or not). This bookkeeping is based on a set of permissions
that regulate the currently allowed mode of access to resources and
closures. It enables us to state and prove a highly informative type
soundness result (see \cref{sec:metatheory}) with expressive invariants that ensure proper
resource usage.

### 2020-05-17 kindly/soa.tex
macro for borrow binding
more comments by Rev#A processed

\paragraph{Substructural type-systems in functional languages}

Many systems propose combinations of
functional programming and linear types in a practical setting.
\lang is to combine key ingredients
from these proposals while still preserving
complete type inference.
Many of the following languages support linear or affine types, but rarely
both. In many cases, it is easy to adapt a system to support both, as
\lang does.
None of the following languages support borrows.

@@

The goal of Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
(LH) is to retrofit linear types to Haskell.
Unlike the previously discussed approaches, LH relies on ``linear
arrows'', written $\multimap$ as in linear logic, 
which are functions that \emph{use} their argument exactly once.
This design is easy to retrofit on top of an existing compiler
such as GHC, but has proven quite controversial\footnote{
  See the in-depth discussion attached to the GHC proposal starting  \href{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}{here}.}.
Most relevant to \lang:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item LH does not admit subtyping for arrows and requires
  $\eta$-expansion to pass unrestricted functions in linear
  contexts. This approach is acceptable in a non-strict language such as
  Haskell but changes the semantics in a strict setting.
\item
  While the LH paper specifies a full type system along with a
  linearity-aware soundness proof, there is neither formal description of
  the type inference algorithm nor a proof of the properties of inference.
  Subsequent work~\cite{DBLP:journals/corr/abs-1911-00268}
  formalizes the inference for rank 1 qualified-types.
  However, there is an implementation of the inference as part of GHC.
\item
  LH promotes a continuation-passing style with functions such as
  \lstinline/withFile : path -> (file ->. file) ->. unit/
  to ensure linear use of resources. This style leads to problems with
  placing the annotation on, e.g., the IO monad.
  \lang follows System F\degree, Quill, and Alms, all of which support
  resource handling in direct style, where types themselves are
  described as affine or linear. (Of course, continuation-passing
  style is also supported.)
  We expect that the direct approach eases modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need to consider the module
  implementation to ensure that linear
  resources are properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

@@

\paragraph{Other substructural type-systems}

@@

\paragraph{Type-system features}
%
\lang relies on constrained types
to introduce the kind inequalities required for linear types.
\hmx~\citep{DBLP:journals/tapos/OderskySW99} 
allows us to use constrained types in an ML-like language with complete
type inference.
\hmx has been shown to be compatible with subtyping,
bounded quantification and existentials~\citep{DBLP:conf/icfp/Simonet03},
GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and there exists a syntactic soundness proof~\citep{DBLP:journals/entcs/SkalkaP02}.
These results make us confident that the system developed in \lang
could be applied to larger and more complex languages such as OCaml
and the full range of features based on ad-hoc polymorphism.
% Alternatively, we could have based \lang on qualified
% types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill.
% This
% choice would also be sustainable as qualified types is part of the
% foundation of Haskell's type system.

>>>>

\subsection{Substructural type-systems in functional languages}

Many systems propose combinations of
functional programming and linear types in a practical setting.
The goal of \lang is to combine key ingredients
from these proposals while still preserving
complete type inference.
Many of the following languages support linear or affine types, but rarely
both. In many cases, it is easy to adapt a system to support both, as
\lang does.
None of the following languages support borrows.

@@

The goal of Linear Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
(LH) is to retrofit linear types to Haskell.
Unlike the previously discussed approaches, LH relies on ``linear
arrows'', written $\multimap$ as in linear logic, 
which are functions that \emph{use} their argument exactly once.
This design is easy to retrofit on top of an existing compiler
such as GHC, but has proven quite controversial\footnote{
  See the in-depth discussion attached to the GHC proposal for LH on GitHub: \url{https://github.com/ghc-proposals/ghc-proposals/pull/111\#issuecomment-403349707}.}.
Most relevant to \lang:
\begin{itemize}[leftmargin=*]
\item LH does not admit subtyping for arrows and requires
  $\eta$-expansion to pass unrestricted functions in linear
  contexts. This approach is acceptable in a non-strict language such as
  Haskell but changes the semantics in a strict setting.
\item
  While the LH paper specifies a full type system along with a
  linearity-aware soundness proof, there is neither formal description of
  the type inference algorithm nor a proof of the properties of inference.
  Subsequent work~\cite{DBLP:journals/corr/abs-1911-00268}
  formalizes the inference for rank 1 qualified-types.
  However, there is an implementation of the inference as part of GHC.
\item
  LH promotes a continuation-passing style with functions such as
  \lstinline/withFile : path -> (file ->. Unrestricted r) ->. r/
  to ensure linear use of resources. This style leads to problems with
  placing the annotation on, e.g., the IO monad.
  \lang follows System F\degree, Quill, and Alms, all of which support
  resource handling in direct style, where types themselves are
  described as affine or linear. (Of course, continuation-passing
  style is also supported.)
  We expect that the direct approach eases modular reasoning about linearity.
  In particular, using abstraction through modules,
  programmers only need to consider the module
  implementation to ensure that linear
  resources are properly handled.
% \item
%   Linear Haskell introduces a notion of linear monad to express
%   imperative code conveniently. Again, this solution is suitable in Haskell,
%   but less appropriate in other contexts. Our borrow system allow
%   to provide a more free-form programming style for imperative code.
% \item
%   Due to the previous remark, ensuring that a given object is never
%   aliased is fairly difficult in Linear Haskell, as one would need to ensure
%   that no non-linear function can manipulate it. On the other
\end{itemize}

@@

\subsection{Other substructural type-systems}

@@

\subsection{Type-system features}
%
\lang relies on constrained types
to introduce the kind inequalities required for linear types.
\hmx~\citep{DBLP:journals/tapos/OderskySW99} 
allows us to use constrained types in an ML-like language with complete
type inference.
\hmx has been shown to be compatible with subtyping,
bounded quantification and existentials~\citep{DBLP:conf/icfp/Simonet03},
GADTs~\citep{DBLP:journals/toplas/SimonetP07},
and there exists a syntactic soundness proof~\citep{DBLP:journals/entcs/SkalkaP02}.
These results make us confident that the system developed in \lang
could be applied to larger and more complex languages such as OCaml
and the full range of features based on ad-hoc polymorphism.
% Alternatively, we could have based \lang on qualified
% types~\cite{DBLP:journals/scp/Jones94}, similarly to Quill.
% This
% choice would also be sustainable as qualified types is part of the
% foundation of Haskell's type system.
